Virtual Memory Assisted Transactional Memory for Dynamic Languages

### 2014-06-02 talk/dls2014/paper/paper.tex
add some refs

In this section, we characterise the model of our TM system and its
guarantees as well as some of the design choices we made. This should
clarify the general semantics in commonly used terms from the
literature.\remi{cite Transactional Memory 2nd edition}

@@

As required for TM systems, we guarantee complete \emph{isolation} and
\emph{atomicity} for transactions at all times. Our method of choice
is \emph{lazy version management}. Modifications by a transaction are
not visible to another transaction before the former commits.
Furthermore, the isolation provides full \emph{opacity} to always
guarantee a consistent read set even for non-committed transactions.
\remi{cite On the Correctness of Transactional Memory}

To also support these properties for irreversible operations that
cannot be undone when we abort a transaction (e.g. I/O, syscalls, and
non-transactional code in general), we use \emph{irrevocable} or
\emph{inevitable transactions}. These transactions are always
guaranteed to commit, which is why they always have to win in case
there is a conflict with another, normal transaction. There is always
at most one such transaction running in the system, thus their
execution is serialised. With this guarantee, providing \emph{strong
isolation} and \emph{serialisability} between non-transactional code
is possible by making the current transaction inevitable right before
running irreversible operations.

@@

In order to reference these objects, we need references that are valid
in all threads and automatically point to the private copies. Since
an object's offset inside a segment is the same in all segments, we
can use this offset to reference objects. Because all segments are
copies of each other, this \emph{Segment Offset ($SO$)} points to the
private version of an object in all threads~/ segments. To then
translate this $SO$ to a real virtual memory address when used inside a
thread, we need to add the thread's segment start address to the
$SO$. The result of this operation is called a \emph{Linear Address
  (LA)}. This is illustrated in Figure \ref{fig:Segment-Addressing}.

@@

In order to eliminate the prohibitive memory requirements of keeping
around $N$ segment copies, we share memory between them. The segments
are initially allocated in a single range of virtual memory by a call
to \lstinline!mmap()!.  As illustrated in Figure
\ref{fig:mmap()-Page-Mapping}, \lstinline!mmap()! creates a mapping
between a range of virtual memory pages and virtual file pages. The
virtual file pages are then mapped lazily by the kernel to real
physical memory pages.

@@

As illustrated in Figure \ref{fig:Page-Remapping}, in our initial
configuration (I) all segments are backed by their own range of
virtual file pages. This is the share-nothing configuration where
all threads have private versions of all objects.

@@

We now use these mechanisms to provide isolation for transactions.
Using write barriers, we implement a \emph{Copy-On-Write (COW)} on the
level of pages. Starting from the initial fully-shared configuration
(Figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
object without other threads seeing the changes immediately, we ensure
that all pages belonging to the object are private to our segment.

@@

\cfbolz{random question: did we investigate the extra memory requirements? we should characterise memory overhead somewhere, eg at least one byte per object for the read markers}

@@

\lstinline!stm_start_transaction()!  starts a transaction. It requires
two arguments, the first being a thread-local data structure and the
second a buffer for use by \lstinline!setjmp()!.
\lstinline!stm_commit_transaction()!  tries to commit the current
transaction. \lstinline!stm_read()!, \lstinline!stm_write()!  perform
a read or a write barrier on an object and \lstinline!stm_allocate()!
allocates a new object with the specified size.
 \lstinline!STM_PUSH_ROOT()!  and \lstinline!STM_POP_ROOT()!  push
and pop objects on the shadow stack~\footnote{A stack for pointers to
  GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.}.  Objects have to be saved using this stack around calls
that may cause a GC cycle to happen, and also while there is no
transaction running. In this simplified API, only
\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
require saving object references.

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
Figure \ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if $SO$ is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

As seen in the API (section~\ref{sub:Application-Programming-Interfac}),
we use a \emph{shadow stack} in order to provide precise garbage
collection.  Any time we call a function that possibly triggers a
collection, we need to save the objects that we need afterwards on the
shadow stack using \lstinline!STM_PUSH_ROOT()!.  That way, they will
not be freed. And in case they were young, we get their new location
in the old object space when getting them back from the stack using
\lstinline!STM_POP_ROOT()!. \remi{cite something which explains
shadowstacks in more detail}

@@

On encountering conflicts, we employ contention management to solve
the problem as well as we can. The general rules are:
\begin{itemize}[noitemsep]
\item prefer transactions that started earlier to younger transactions
\item to support \emph{inevitable} transactions, we always prefer them
  to others since they cannot abort
\end{itemize}
We can either simply abort a transaction to let the other one succeed,
or we can also wait until the other transaction committed. The latter
is an interesting option if we are trying to commit a write and
another transaction already read the object. We can then signal the
other transaction to commit as soon as possible and wait. After
waiting, there is now no conflict between our write and the already
committed read anymore.

@@

As expected, all interpreters with a GIL do not scale with the number
of threads. They even become slower because of the overhead of
thread-switching and GIL handling. We also see Jython scale when we
expect it to (mandelbrot, raytrace, richards), and behave similar to
the GIL interpreters in the other cases.

@@

@@

>>>>

In this section, we characterise the model of our TM system and its
guarantees as well as some of the design choices we made. This should
clarify the general semantics in commonly used terms from the
literature~\cite{harris10}.

@@

As required for TM systems, we guarantee complete \emph{isolation} and
\emph{atomicity} for transactions at all times. Our method of choice
is \emph{lazy version management}. Modifications by a transaction are
not visible to another transaction before the former commits.
Furthermore, the isolation provides full
\emph{opacity}~\cite{guerraoui08} to always guarantee a consistent
read set even for non-committed transactions.

To also support these properties for irreversible operations that
cannot be undone when we abort a transaction (e.g. I/O, syscalls, and
non-transactional code in general), we use \emph{irrevocable} or
\emph{inevitable transactions}~\cite{blundell06,spear08}. These transactions are always
guaranteed to commit, which is why they always have to win in case
there is a conflict with another, normal transaction. There is always
at most one such transaction running in the system, thus their
execution is serialised. With this guarantee, providing \emph{strong
isolation} and \emph{serialisability} between non-transactional code
is possible by making the current transaction inevitable right before
running irreversible operations.

@@

In order to reference these objects, we need references that are valid
in all threads and automatically point to the private copies. Since
an object's offset inside a segment is the same in all segments, we
can use this offset to reference objects. Because all segments are
copies of each other, this \emph{Segment Offset ($SO$)} points to the
private version of an object in all threads~/ segments. To then
translate this $SO$ to a real virtual memory address when used inside a
thread, we need to add the thread's segment start address to the
$SO$. The result of this operation is called a \emph{Linear Address
  (LA)}. This is illustrated in figure \ref{fig:Segment-Addressing}.

@@

In order to eliminate the prohibitive memory requirements of keeping
around $N$ segment copies, we share memory between them. The segments
are initially allocated in a single range of virtual memory by a call
to \lstinline!mmap()!.  As illustrated in figure
\ref{fig:mmap()-Page-Mapping}, \lstinline!mmap()! creates a mapping
between a range of virtual memory pages and virtual file pages. The
virtual file pages are then mapped lazily by the kernel to real
physical memory pages.

@@

As illustrated in figure \ref{fig:Page-Remapping}, in our initial
configuration (I) all segments are backed by their own range of
virtual file pages. This is the share-nothing configuration where
all threads have private versions of all objects.

@@

We now use these mechanisms to provide isolation for transactions.
Using write barriers, we implement a \emph{Copy-On-Write (COW)} on the
level of pages. Starting from the initial fully-shared configuration
(figure \ref{fig:Page-Remapping}, (II)), when we need to modify an
object without other threads seeing the changes immediately, we ensure
that all pages belonging to the object are private to our segment.

@@

@@

\lstinline!stm_start_transaction()!  starts a transaction. It requires
two arguments, the first being a thread-local data structure and the
second a buffer for use by \lstinline!setjmp()!.
\lstinline!stm_commit_transaction()!  tries to commit the current
transaction. \lstinline!stm_read()!, \lstinline!stm_write()!  perform
a read or a write barrier on an object and \lstinline!stm_allocate()!
allocates a new object with the specified size.
 \lstinline!STM_PUSH_ROOT()!  and \lstinline!STM_POP_ROOT()!  push
and pop objects on the shadow stack~\footnote{A stack for pointers to
  GC objects that allows for precise garbage collection. All objects
  on that stack are never seen as garbage and are thus always kept
  alive.~\cite{fergus02}}.  Objects have to be saved using this stack around calls
that may cause a GC cycle to happen, and also while there is no
transaction running. In this simplified API, only
\lstinline!stm_allocate()!  and \lstinline!stm_commit_transaction()!
require saving object references.

@@

However, the layout of a segment is not uniform and we actually
privatise a few areas again right away. These areas are illustrated in
figure \ref{fig:Segment-Layout} and explained here:
\begin{description}[noitemsep]
\item [{NULL~page:}] This page is unmapped and will produce a
  segmentation violation when accessed. We use this to detect
  erroneous dereferencing of \lstinline!NULL! references.  All
  $\%gs{::}SO$ translated to linear addresses will point to NULL pages
  if $SO$ is set to \lstinline!NULL!.
\item [{Segment-local~data:}] Some area private to the segment that
  contains segment-local information.
\item [{Read~markers:}] These are pages that store information about
  which objects were read in the current transaction running in this
  segment.
\item [{Nursery:}] This area contains all the freshly allocated
  objects (\emph{young objects}) of the current transaction. The GC
  uses pointer-bump allocation in this area to allocate objects in the
  first generation.
\item [{Old~object~space:}] These pages are the ones that are really
  shared between segments. They mostly contain old objects but also
  some young ones that were too big to be allocated in the nursery.
\end{description}

@@

As seen in the API (section~\ref{sub:Application-Programming-Interfac}),
we use a \emph{shadow stack}~\cite{fergus02} in order to provide precise garbage
collection.  Any time we call a function that possibly triggers a
collection, we need to save the objects that we need afterwards on the
shadow stack using \lstinline!STM_PUSH_ROOT()!.  That way, they will
not be freed. And in case they were young, we get their new location
in the old object space when getting them back from the stack using
\lstinline!STM_POP_ROOT()!.

@@

On encountering conflicts, we employ contention management to solve
the problem as well as we can. The general rules are:
\begin{itemize}[noitemsep]
\item prefer transactions that started earlier to younger transactions
\item to support \emph{inevitable} transactions, we always prefer them
  to others since they cannot abort (similar to \cite{blundell06})
\end{itemize}
We can either simply abort a transaction to let the other one succeed,
or we can also wait until the other transaction committed. The latter
is an interesting option if we are trying to commit a write and
another transaction already read the object. We can then signal the
other transaction to commit as soon as possible and wait. After
waiting, there is now no conflict between our write and the already
committed read anymore.

@@

As expected, all interpreters with a GIL do not scale with the number
of threads. They even become slower because of the overhead of
thread-switching and GIL handling (see \cite{beazley10} for a detailed
analysis). We also see Jython scale when we expect it to (mandelbrot,
raytrace, richards), and behave similar to the GIL interpreters in the
other cases.

@@

Eliminate GIL:
\begin{itemize}
\item Previous attempts with HTM: \cite{nicholas06,odaira14,fuad10}
\item Previous attempts with STM: \cite{stmupdate13}
\end{itemize}

Similar STMs:
\begin{itemize}
\item FastLane: \cite{warmhoff13}
\item TML: \cite{spear09}
\item Virtualizing HTM: \cite{rajwar05}
\item Page-based virtualizing HyTM: \cite{chung06} (XTM can be
  implemented either in the OS as part of the virtual memory manager or
  between underlying TM systems and the OS, like virtual machines;
  Conflicts for overflowed transactions are tracked at page granularity;
  XTM-e allows conflict detection at cache line granu-
  larity, even for overflowed data in virtual memory)
\item using mmap(): Memory-Mapped Transactions
\item mem-protected conflict detection: \cite{martin09}
\end{itemize}

@@

\bibitem{beazley10} Beazley, David. "Understanding the python gil."
  \emph{PyCON Python Conference}. Atlanta, Georgia. 2010.

\bibitem{harris10} Harris, Tim, James Larus, and Ravi
  Rajwar. "Transactional memory." \emph{Synthesis Lectures on Computer
  Architecture 5.1} (2010): 1-263.

\bibitem{guerraoui08} Guerraoui, Rachid, and Michal Kapalka. "On the
  correctness of transactional memory." \emph{Proceedings of the 13th
    ACM SIGPLAN Symposium on Principles and practice of parallel
    programming.} ACM, 2008.

\bibitem{blundell06} Blundell, Colin, E. Christopher Lewis, and Milo
  Martin. "Unrestricted transactional memory: Supporting I/O and system
  calls within transactions." (2006).
\bibitem{spear08} Spear, Michael F., et al. "Implementing and
  exploiting inevitability in software transactional memory."
  \emph{Parallel Processing, 2008}. ICPP'08. 37th International
  Conference on. IEEE, 2008.

\bibitem{fergus02} Fergus Henderson. 2002. Accurate garbage collection
  in an uncooperative environment. \emph{In Proceedings of the 3rd
    international symposium on Memory management} (ISMM '02).

\bibitem{stmupdate13} Armin Rigo, Remigius Meier. Update on
  STM. \url{morepypy.blogspot.ch/2013/10/update-on-stm.html}

\bibitem{rajwar05} Rajwar, Ravi, Maurice Herlihy, and Konrad
  Lai. "Virtualizing transactional memory." \emph{Computer
    Architecture}, 2005. ISCA'05. Proceedings. 32nd International
  Symposium on. IEEE, 2005.

\bibitem{chung06} Chung, JaeWoong, et al. "Tradeoffs in transactional
  memory virtualization." \emph{ACM SIGARCH Computer Architecture
  News}. Vol. 34. No. 5. ACM, 2006.

\bibitem{martin09} Martín Abadi, Tim Harris, and Mojtaba
  Mehrara. 2009. Transactional memory with strong atomicity using
  off-the-shelf memory protection hardware. SIGPLAN Not. 44, 4 (February
  2009), 185-196.
